Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-14 talk/icooolps2011/code/map.tex
use the (admittedly obscure) name "trace-elidable", simply because nobody has preconceptions abous it

    @invariant
    def getindex(self, name):
        return self.indexes.get(name, -1)

    @invariant
    def add_attribute(self, name):
        if name not in self.other_maps:
            newmap = Map()
            newmap.indexes.update(self.indexes)
            newmap.indexes[name] = len(self.indexes)
            self.other_maps[name] = newmap
        return self.other_maps[name]

>>>>

    @elidable
    def getindex(self, name):
        return self.indexes.get(name, -1)

    @elidable
    def add_attribute(self, name):
        if name not in self.other_maps:
            newmap = Map()
            newmap.indexes.update(self.indexes)
            newmap.indexes[name] = len(self.indexes)
            self.other_maps[name] = newmap
        return self.other_maps[name]

### 2011-04-14 talk/icooolps2011/paper.tex
use the (admittedly obscure) name "trace-elidable", simply because nobody has preconceptions abous it

\cfbolz{XXX stress more that "the crux of the techniques and a significant
portion of new contributions in the paper are from how to refactoring codes to
expose likely runtime constants and invariant functions"}

@@

This hint can be used to mark functions as \emph{invariant}. A function is
termed invariant if, during the execution of the program, successive calls to
the function with identical arguments always return the same result. From this
definition follows that a call to an invariant function with constant arguments
in a trace can be replaced with the result of the call.

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is an invariant function. To communicate this, there is a \texttt{invariant} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    @invariant
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

@@

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as invariant, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

@@

On the one hand, the \texttt{invariant} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
function is annotated to be invariant, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{invariant}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}. We hope to introduce a
debugging mode which would (slowly) check whether the annotation is applied
incorrectly to mitigate this problem.

\subsubsection{Observably invariant Functions}

@@

Why can't we simply write an analysis to find out that the \texttt{x} fields of the
\texttt{A} instances is immutable and deduce that \texttt{compute} is an invariant function,
since it only reads the \texttt{x} field and does not have side effects? This might
be possible in this particular case, but in practice the functions that are
annotated with the \texttt{invariant} decorator are usually more complex.
The easiest example for this is that of a function that uses memoization to
cache its results. If this function is analyzed, it looks like the function has
side effects, because it changes the memoizing dictionary. However, because this side
effect is not externally visible, the function is still invariant. This is
a property that is not easily detectable by analysis.

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints we have looked at in the
two previous sections don't seem to help with the current object model. There is
no invariant function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

This is a common problem when trying to apply hints. Often, the interpreter
needs a small rewrite to expose the invariant functions and nearly-constant objects
that are implicitly there. In the case of instance fields this rewrite is not
entirely obvious. The basic idea is as follows. In theory instances can have
arbitrary fields. In practice however many instances share their layout (i.e.
their set of keys) with many other instances.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Therefore they have to be immutable, which means
that their \texttt{getindex} method is an invariant function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also invariant,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

@@

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
an invariant function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

@@

What we would really like is if the \texttt{Class.find\_method} method were invariant.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is an invariant operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@invariant}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
invariant, because when the version object changes, the result of the call might be
different than the previous one.

@@

%\subsection{More General Patterns}
%
%The techniques we used above to make instance and class lookups faster are
%applicable in more general cases than the one we developed them for. A more
%abstract view of maps is that of splitting a data-structure into a part that
%changes slowly, and a part that changes quickly. In the concrete example of maps
%we split the original dictionary into the map (the slow-changing part) and the
%storage array (the quick-changing part). All the computation on the
%slow-changing part can be constant-folded during tracing so that only the
%manipulation of the quick-changing part remains.
%
%Similarly, versions can be used to constant-fold arbitrary functions of large data
%structures. The version needs to be updated carefully every time the result of
%this function can change. Therefore this is useful only if the data structure is
%expected to change slowly.

>>>>

\cfbolz{XXX stress more that "the crux of the techniques and a significant
portion of new contributions in the paper are from how to refactoring codes to
expose likely runtime constants and trace-elidable functions"}

@@

XXX not too happy with the definition

This hint can be used to mark functions as \emph{trace-elidable}. A function is
termed trace-elidable if, during the execution of the program, the results of
subsequent calls to the function with identical arguments may be be replaced
with the result of the first call without changing the program's behaviour.
From this
definition follows that a call to an trace-elidable function with constant arguments
in a trace can be replaced with the result of the call.\footnote{This property
is less strict than that of a "pure" function, because it is only about actual
calls during execution. All pure functions are trace-elidable though.}

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    @elidable
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

@@

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

@@

On the one hand, the \texttt{elidable} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
function is annotated to be trace-elidable, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{elidable}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}. We hope to introduce a
debugging mode which would (slowly) check whether the annotation is applied
incorrectly to mitigate this problem.

\subsubsection{Observably trace-elidable Functions}

@@

Why can't we simply write an analysis to find out that the \texttt{x} fields of the
\texttt{A} instances is immutable and deduce that \texttt{compute} is a trace-elidable function,
since it only reads the \texttt{x} field and does not have side effects? This might
be possible in this particular case, but in practice the functions that are
annotated with the \texttt{elidable} decorator are usually more complex.
The easiest example for this is that of a function that uses memoization to
cache its results. If this function is analyzed, it looks like the function has
side effects, because it changes the memoizing dictionary. However, because this side
effect is not externally visible, the function is still trace-elidable. This is
a property that is not easily detectable by analysis.

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints we have looked at in the
two previous sections don't seem to help with the current object model. There is
no trace-elidable function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

This is a common problem when trying to apply hints. Often, the interpreter
needs a small rewrite to expose the trace-elidable functions and nearly-constant objects
that are implicitly there. In the case of instance fields this rewrite is not
entirely obvious. The basic idea is as follows. In theory instances can have
arbitrary fields. In practice however many instances share their layout (i.e.
their set of keys) with many other instances.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Therefore they have to be immutable, which means
that their \texttt{getindex} method is an trace-elidable function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

@@

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a trace-elidable function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

@@

What we would really like is if the \texttt{Class.find\_method} method were trace-elidable.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a trace-elidable operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@elidable}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
trace-elidable, because when the version object changes, the result of the call might be
different than the previous one.

@@

%\subsection{More General Patterns}
%
%The techniques we used above to make instance and class lookups faster are
%applicable in more general cases than the one we developed them for. A more
%abstract view of maps is that of splitting a data-structure into an immutable part (\eg the map)
%and a part that changes (\eg the storage array). All the computation on the
%immutable part is trace-elidable so that only the manipulation of the quick-changing
%part remains in the trace after optimization.
%
%Similarly, versions can be used to constant-fold arbitrary functions of large data
%structures. The version needs to be updated carefully every time the result of
%this function can change. Therefore this is useful only if the data structure is
%expected to change slowly.
